Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

### 2017-10-12 experiments.tex
Edited experiments too

\subsection{FINDING GENERATING STRUCTURE}

@@

\subsection{PREDICTION AND PARSIMONY}

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
for all of them. In each experiment, random 10\% of the original
data set was selected as training data and the rest 90\% as test data. The training was conducted using the same dynamic programming based exact structure learning algorithm.

When predicting with structures learned by the BDeu score, we used the
Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter-free, for structures learned by the other model
selection criteria, we used the sequential predictive NML (sNML)
parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

\begin{table}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different data sets.}
\label{tbl:preds}
\begin{center}
\begin{tabular}{crrrrr}
       Data
    & N
    & \multicolumn{1}{p{0.7cm}}{\centering BDeu \\ BPP}
    & \multicolumn{1}{p{0.8cm}}{\centering BIC \\ sNML}
    & \multicolumn{1}{p{0.9cm}}{\centering fNML \\ sNML}
    & \multicolumn{1}{p{0.9cm}}{\centering qNML \\ sNML}\\
       \midrule
    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
 PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
 Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
 HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
 BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
 HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
 HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
 BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
 Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
 TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
 Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
 Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
 PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
 Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
\end{tabular}
\end{center}
\end{table}

Table~\ref{tbl:preds} features the average predictive losses for different
data sets.  BIC's bias for simplicity makes it often win (written bold
in the table) with small sample sizes, but it performs worst (in
italics) for the large sample sizes. the qNML seems to achieve a lot
of runner-up positions (underlined) making it a rather safe choice.

Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
strangely in terms of model complexity here measured by the number of
parameters in the model. qNML, instead, appears to yield more
parsimonious models.

Looking at the number of parents for the same 20 data sets again
features BIC's preference for simple models
(Table~\ref{tbl:nofparams}).  qNML usually (18/20) yields more
parsimonious models than fNML that selects the most complex model for 14
out of 20 data sets.

\begin{table}
  \caption{Average number of parents per node 
    for different model selection criteria in 20 different data sets.}
\label{tbl:nofparams}
\begin{center}
\begin{tabular}{crrrrr}
%%    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
       Data
    & N
    & \multicolumn{1}{p{0.7cm}}{\centering BDeu\\ }
    & \multicolumn{1}{p{0.8cm}}{\centering BIC\\ }
    & \multicolumn{1}{p{0.9cm}}{\centering fNML\\}
    & \multicolumn{1}{p{0.9cm}}{\centering qNML\\}\\
\midrule
    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
 PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
 Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
 HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
 BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
 HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
 HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
 BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
 Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
 TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
 Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
 Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
 PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
 Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
\end{tabular}
\end{center}
\end{table}

>>>>

\subsection{Finding Generating Structure}

@@

\subsection{Prediction and Parsimony}

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and ran train and test experiments for
all of them. To better compare the performance over different sample
sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
the data sets as training data and the rest as the test data.
This was done for 1000 different permutations of each dataset.
The training was conducted using the dynamic
programming based exact structure learning algorithm.

When predicting $P(d_{test}|D_{train},S,\theta)$,
with structures $S$ learned by the BDeu score, we used the
Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
hyperparameter-free, for structures learned by the other model
selection criteria, we used the sequential predictive NML (sNML)
parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.

\begin{table}
\caption{Average predictive performance rank over different sample sizes for different model selection criteria in 20 different data sets.}
\label{tbl:preds}
\begin{center}
\begin{tabular}{crrrrr}
       Data
    & N
    & \multicolumn{1}{p{0.7cm}}{\centering BDeu \\ BPP}
    & \multicolumn{1}{p{0.8cm}}{\centering BIC \\ sNML}
    & \multicolumn{1}{p{0.9cm}}{\centering fNML \\ sNML}
    & \multicolumn{1}{p{0.9cm}}{\centering qNML \\ sNML}\\
       \midrule
 PostOpe &     90 &              2.79 &     \textbf{1.20} &  \underline{3.06} &           2.94 \\
    Iris &    150 &  \underline{2.82} &              2.37 &     \textbf{2.27} &           2.54 \\
    Wine &    178 &  \underline{3.23} &     \textbf{1.88} &              2.67 &           2.22 \\
   Glass &    214 &  \underline{3.61} &              3.09 &     \textbf{1.42} &           1.88 \\
 Thyroid &    215 &              2.55 &  \underline{3.21} &     \textbf{1.80} &           2.44 \\
 HeartSt &    270 &  \underline{3.12} &     \textbf{1.39} &              3.12 &           2.37 \\
 BreastC &    286 &  \underline{3.09} &     \textbf{1.41} &              2.97 &           2.53 \\
 HeartHu &    294 &  \underline{3.18} &     \textbf{1.66} &              2.90 &           2.27 \\
 HeartCl &    303 &  \underline{3.46} &     \textbf{1.38} &              2.99 &           2.17 \\
   Ecoli &    336 &              3.20 &  \underline{3.53} &     \textbf{1.24} &           2.04 \\
   Liver &    345 &  \underline{3.17} &              2.39 &              2.69 &  \textbf{1.75} \\
 Balance &    625 &  \underline{3.35} &              1.91 &     \textbf{1.59} &           3.16 \\
 BcWisco &    699 &  \underline{3.06} &              2.03 &              2.89 &  \textbf{2.02} \\
 Diabete &    768 &  \underline{2.91} &              2.70 &              2.68 &  \textbf{1.71} \\
 TicTacT &    958 &  \underline{3.44} &              2.71 &     \textbf{1.31} &           2.53 \\
   Yeast &   1484 &              2.60 &  \underline{3.76} &     \textbf{1.55} &           2.10 \\
 Abalone &   4177 &              2.60 &  \underline{3.64} &     \textbf{1.04} &           2.72 \\
 PageBlo &   5473 &              2.24 &  \underline{3.61} &     \textbf{1.31} &           2.83 \\
   Adult &  32561 &              3.23 &  \underline{3.77} &     \textbf{1.00} &           2.00 \\
 Shuttle &  58000 &     \textbf{1.44} &  \underline{3.78} &              1.56 &           3.22 \\
\end{tabular}
\end{center}
\end{table}

For each train/test sample, we ranked the predictive performance of
the models learned by different four structure learning criteria (rank
1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
average rank for different data sets, the average being taken over
1000 different train test samples for each 9 sample sizes for each
data.  BIC's bias for simplicity makes it often win (written bold in
the table) with small sample sizes, but it often performs worst
(underlined) for the larger sample sizes (for the same reason), while
fNML seems to be good for large sample sizes. The striking feature
about the qNML is its robustness.  It is usually between BIC and fNML
for all the sample sizes making it a ``safe choice''. This can be
quantified if we further average the columns of the
table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
$0.43$.  While fNML achieves on average the best rank, the runner-up
qNML has the lowest standard deviation.

Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
strangely in terms of model complexity here measured by the number of
parameters in the model. qNML, instead, appears to yield more
parsimonious models. To study the concern of fNML producing too
complex models for small sample sizes, we studied the number of
parameters in models produced by different model selection criteria
when using 10\% of each data set for structure learning.

Looking at the number of parameters for the same 20 data sets again
features BIC's preference for simple models
(Table~\ref{tbl:nofparams}).  qNML usually (19/20) yields more
parsimonious models than fNML that selects the most complex model for
7 out of 20 data sets.

The graphs for different sample sizes for both predictive accuracy and
the number of parameters can be found in the Supplementary Material.

\begin{table}
  \caption{Average number of parameters in models 
    for different model selection criteria in 20 different data sets.}
\label{tbl:nofparams}
\begin{center}
\begin{tabular}{crrrrr}
%%    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
       Data
    & N
    & \multicolumn{1}{p{0.7cm}}{\centering BDeu\\ }
    & \multicolumn{1}{p{0.8cm}}{\centering BIC\\ }
    & \multicolumn{1}{p{0.9cm}}{\centering fNML\\}
    & \multicolumn{1}{p{0.9cm}}{\centering qNML\\}\\
\midrule
    Iris &    15 &     \underline{37} &   \textbf{23} &                33 &               29 \\
 PostOpe &    18 &   \underline{1217} &   \textbf{19} &               397 &              146 \\
   Ecoli &    34 &    \underline{182} &   \textbf{31} &               162 &               77 \\
   Liver &    35 &                 45 &   \textbf{15} &    \underline{61} &               24 \\
    Wine &    36 &  \underline{16521} &   \textbf{70} &               807 &              205 \\
   Glass &    44 &   \underline{1677} &   \textbf{48} &               506 &               97 \\
 Thyroid &    44 &                 40 &   \textbf{23} &    \underline{66} &               28 \\
 HeartSt &    54 &  \underline{16861} &   \textbf{44} &              1110 &              256 \\
 BreastC &    58 &  \underline{25797} &   \textbf{49} &              3767 &              844 \\
 HeartHu &    60 &   \underline{1634} &   \textbf{43} &               792 &               90 \\
 HeartCl &    62 &  \underline{34381} &   \textbf{47} &              1433 &              404 \\
 BcWisco &    70 &   \underline{4630} &   \textbf{42} &               603 &               89 \\
 Diabete &    77 &                 39 &   \textbf{22} &   \underline{216} &               34 \\
 TicTacT &    96 &  \underline{13701} &   \textbf{25} &              1969 &              767 \\
 Balance &   126 &        \textbf{20} &            24 &                49 &  \underline{611} \\
   Yeast &   149 &                 71 &   \textbf{31} &   \underline{265} &               75 \\
 Abalone &   418 &                 91 &   \textbf{46} &   \underline{150} &               63 \\
 PageBlo &   548 &    \underline{703} &   \textbf{45} &               380 &               56 \\
 Shuttle &  5800 &                535 &   \textbf{99} &   \underline{717} &              130 \\
   Adult &  6513 &                699 &  \textbf{479} &  \underline{1555} &              945 \\
\end{tabular}
\end{center}
\end{table}

### 2017-10-12 qnml.tex
Edited experiments too

\subsection {qNML IS SCORE EQUIVALENT}

@@

\subsection{qNML IS CONSISTENT}

@@

\subsection{qNML EQUALS NML FOR MANY MODELS}
The fNML criterion can be seen as a computationally feasible
approximation of the more desirable NML criterion.  However, the fNML
criterion equals the NML criterion only for the Bayesian network
structure with no arcs.  We will next show that the qNML criterion
equals the NML criterion for all the networks $G$ whose connected
components tournaments (i.e., complete directed acyclic subgraphs of
$G$).

\begin{theorem}
If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
for all data sets $D$.
\end{theorem}
\begin{proof}
We first show that the NML-criterion for a Bayesian network
decomposes by the connected components.

Because the maximum likelihood for the data $D$ decomposes,
we can write
\begin{eqnarray}
  \lefteqn{P_{NML}(D;G)} \nonumber \\
  && = \frac{P(D;\hat\theta(D),G)}
            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
\end{eqnarray}

Clearly, the qNML score also decomposes by the connected components,
so it remains to show that if the (sub)network $G$ is a tournament,
then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
score equivalence of the NML criterion and the qNML criterion, we may
pick a tournament $G$ such that the linear ordering defined by $G$
matches the ordering of the data columns, i.e., $i<j$ implies $G_i
\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
criterion we see that for the tournament $G$, the sum telescopes
leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
follows, since for any data vector $x$ in data $D$, we have
$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$, where $P^1$ denotes
the model that takes $n$-dimensional vectors to be values of the
single (collapsed) categorical variable.  Denoting prefixes of data
vector $x$ by $x^{:i}$, and the number of times such a prefix appears
on $N$ rows $[d_1,\ldots,d_N]$ of the $N\times n$ data matrix $D$ by
$N_D(x^{:i})$, (so that $N_D(x^{:0})=N)$, we have
\begin{eqnarray}
  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{N_D(d_j^{:i})}{N_D(d_j^{:i-1})} \nonumber\\
  &=& \prod_{j=1}^{N} \frac{N_D(d_j^{:n})}{N} \nonumber \\
  &=& P^1(D;\hat\theta^1(D),G).
\end{eqnarray}
Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
maximum likelihood probabilities, the equality above implies the equality
of these distributions.
\end{proof}

Equality established, we would like to still state the number $a(n)$
of different $n$-node networks whose connected components are
tournaments.  We start by generating all the $p(n)$ integer partitions
i.e. ways to partition $n$ labelled into parts with different
size-profiles.  For example, for $n=4$, we have $p(4)=5$ partition
profiles $[[4], [3, 1], [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of
these partition size profiles corresponds to many different networks,
apart from the last one that corresponds just to the empty network.
We count number of networks for one such partition size-profile (and
then later sum these counts up).  For any such partition profile
$(p_1,\ldots,p_k)$ we can count the ways we can assign the nodes to
different parts and then order each part. This leads to the product
$n\choose p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
the order of different parts of the same size does not matter, so for
all groups of parts having the same size, we have to divide the
product above by the factorial of the size of such group. Notice also,
that the product above telescopes, leaving us a formula for OEIS
sequence A000262\footnote{https://oeis.org/A000108} as
described by Thomas Wieder:

\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
number of different parts of the $i^\text{th}$ partition of $n$,
$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
partition of $n$, one has:}
$$
a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
$$
For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.

\subsection{QNML IS REGULAR}

>>>>

\subsection {qNML Is Score Equivalent}

@@

\subsection{qNML Is Consistent}

@@

\subsection{qNML Equals NML for Many Models}
The fNML criterion can be seen as a computationally feasible
approximation of the more desirable NML criterion.  However, the fNML
criterion equals the NML criterion only for the Bayesian network
structure with no arcs.  It can be shown that the qNML criterion
equals the NML criterion for all the networks $G$ whose connected
components are tournaments (i.e., complete directed acyclic subgraphs of
$G$). These networks include the empty network, the fully connected
one and many networks in between having different complexity. While
the generating network is unlikely to be composed of tournament
components, the result increases the plausibility that qNML is a
reasonable approximation for NML in general\footnote{A claim that is
  naturally subject for further study.}.

\begin{theorem}
If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
for all data sets $D$.
\end{theorem}
\begin{proof}
The full proof can be found in Appendix B in the Supplementary
Material.  The proof first shows that NML decomposes for these
particular structures, so it is enough to show the equivalence for
fully connected graphs.
It further derives the number $a(n)$ of
different $n$-node networks whose connected components are
tournaments, which turns out to be the formula for OEIS sequence
A000262\footnote{https://oeis.org/A000108}.
In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051,
37633, 394353, 4596553, \ldots$.
\end{proof}

\subsection{qNML Is Regular}
